A minimal reaction-diffusion neural model generates {\emph{C. elegans}} undulation

### 2020-12-27 paper.tex
fix some small problems
Changes from paper review

    The small (1 mm) nematode \emph{Caenorhabditis elegans} \cite{corsi2015} has become widely used as a model organism; in particular the \emph{C. elegans} connectome has been completely mapped, and \emph{C. elegans} locomotion has been widely studied (\citet{corsi2015}). We describe a minimal reaction-diffusion model for the locomotion of \emph{C. elegans}, using as a framework a simplified version of Xu et al.'s "descending pathway" description of the \emph{C. elegans} central pattern generator (CPG) \citet{xu2018,wen2012}.  Finally, we realize a model of the required oscillations and coupling with a network of coupled \citet{keener1983} analog neurons.
    We note that \citet{olivares2019} present a likely more realistic model which relies on small networks of neurons, and presents a distributed model of the CPG. We use the simpler simulation to show that a small network of \citet{fitzhugh1955}-\citet{nagumo1962} neurons (one of the simplest neuronal models) can generate key features of \emph{C. elegans} undulation (see \citet{magnes2012}), and thus locomotion.
\end{abstract}

@@

There are many other models for the CPG; for example, a 2019 preprint by \cite{olivares2019} proposes a distributed network of self-oscillating systems of neurons, instead of a structured chain like our proposed CPG.  The point of this paper is to create a minimal working model, though, and not to strive for full realism.  Our CPG model can be interpreted as having a graph-like structure, and it is easy to manipulate the "weights" of the edges (connections between neurons) of this "graph" to study defects or other changes in the CPG in an efficient, reproducible and explainable way.

@@

In more complex animals, like horses or dogs, there is only one CPG; the dynamics of the CPG change depending on the gait of the animal \citep{collins1994}.  The mode of locomotion depends on the sequence in which the neurons fire, but the network's topology remains the same.  We show later that breaking the symmetry of the CPG model can lead to radical changes in behaviour.

\begin{figure}[h!]
    % begin a continued float; see 
    % https://tex.stackexchange.com/questions/118192/figure-subnumbering-1a
    \ContinuedFloat* 
    \centering
    \includegraphics[width=9cm]{figures/xu_cpg/xu_cpg.png}
    \caption{The central pattern generator proposed by \citet{xu2018}.  Note the structure of the neurons, especially the head oscillator driving the descending pathway.}
    \label{fig: xu_cpg}
\end{figure}

@@

\begin{equation}
    \label{eq: fhn}
    \begin{aligned}
        \frac{dv}{dt} &= f(v) + w + I_\mathrm{ext}\\
        \frac{dw}{dt} &= \varepsilon(v - \gamma w + \alpha )\\
        f(v) &= \frac{v^3}{3} - v
    \end{aligned}
\end{equation}

@@

% remark that if the w and v nullclines come near each other, you see a critical slowdown.
% In Hodgkin-Huxley notation, $n$ is the potassium channel.  In the FHN formulation, this corresponds to the $w$ gate variable.
\begin{figure}[h!]
    \label{fig: nm nullclines}
    \centering
    \includegraphics{figures/neuron_dynamics/neuron_dynamics.pdf}
    % note that in the vicinity of the v-nullcline, the w-nullcline is essentially straight.
    \caption{Dynamics of \textit{C. elegans} neurons following Xu \textit{et al.} \cite{xu2018} (left )and Fitzhugh-Nagumo neurons \cite{fitzhugh1955, nagumo1962} (right).}
\end{figure}

@@

\begin{equation}
    \label{eq: fhnd}
    \begin{aligned}
        \frac{dv}{dt}   &= f(v) + w  + \color{green}{D(Δv)}\\
        \frac{dw}{dt}   &= ϵ(a - bv)\\
        f(v) &= \frac{v^3}{3} - v
    \end{aligned}
\end{equation}

@@

We performed the simulations in Python, using the standard SciPy ODE solvers, which wrap LSODA. % TODO list solver here

The main function of the code calculates the left-hand side of Eq. 1 of the system of differential equations which is materialized from our CPG connectome, when given the state and a set of unchanging parameters.  This procedure is followed by the ODE solver which solves the trajectory of the system given some initial conditions; for example, if the neuron does not initially fire, the system may remain in a steady state and not oscillate.

\subsection{Methods}

@@

\subsection{Generating undulation from neuron potential}

@@

@@

\begin{figure}[h!]
    \centering
    \pgfplotsset{}
    % \inputtikz{figures/anal_sim_comp/anal_sim_comp}
    \includegraphics[height=9cm]{figures/anal_sim_comp/anal_sim_comp.pdf}
    \caption{Comparison between computationally simulated dynamics (top row) and observed dynamics in the analog circuit (bottom row).  (a) is the nullcline for the FitzHugh-Nagumo (simulated) equations, (b) is the "head oscillator" simulated (two neurons which mutually inhibit each other), (c) is the "descending pathway" simulated (one neuron driving another).  (d) is the nullcline for the analog circuit - a piecewise linear approximation to the cubic, (e) is the analog version of the head oscillator, and (f) is the analog version of the descending pathway.}
    \label{fig: anal_sim_comp}
\end{figure}

>>>>

    The small (1 mm) nematode \emph{Caenorhabditis elegans} (see wormbook.org) has become widely used as a model organism; in particular the \emph{C. elegans} connectome has been completely mapped, and \emph{C. elegans} locomotion has been widely studied (\citet{corsi2015}). We describe a minimal reaction-diffusion model for the locomotion of \emph{C. elegans}, using as a framework a simplified version of Xu et al.'s "descending pathway" description of the \emph{C. elegans} central pattern generator (CPG) \citet{xu2018,wen2012}.  Finally, we realize a model of the required oscillations and coupling with a network of coupled \citet{keener1983} analog neurons.
    We note that \citet{olivares2019} present a likely more realistic model which relies on small networks of neurons, and presents a distributed model of the CPG. We use the simpler simulation to show that a small network of \citet{fitzhugh1955}-\citet{nagumo1962} neurons (one of the simplest neuronal models) can generate key features of \emph{C. elegans} undulation (see \citet{magnes2012}), and thus locomotion.
\end{abstract}

@@

There are many other models for the CPG; for example, a 2019 preprint by \cite{olivares2019} proposes a distributed network of self-oscillating systems of neurons, instead of a structured chain like our proposed CPG.  The point of this paper is to create a minimal working model, though, and not to strive for full realism.  Our CPG model can be interpreted as having a graph-like structure, and it is easy to manipulate the "weights" of the edges (connections between neurons) of this "graph" to study defects or other changes in the CPG in an efficient, reproducible and explainable way.  Additionally, the simplicity of the model means that it is easy, given a set of parameters, to investigate and reason about the reason for some behaviour of a worm.  Since conditions in the lab are not exactly controllable, our model allows for the

@@

In more complex animals, like horses or dogs, there is only one CPG; the dynamics of the CPG change depending on the gait of the animal \citep{collins1994}.  The mode of locomotion depends on the sequence in which the neurons fire, but the network's topology remains the same.  We show later that breaking the symmetry of the CPG model can lead to radical changes in behaviour.
\setcounter{figure}{1}
\begin{figure}[h!]
    % begin a continued float; see 
    % https://tex.stackexchange.com/questions/118192/figure-subnumbering-1a
    \ContinuedFloat* 
    \centering
    \includegraphics[width=9cm]{figures/xu_cpg/xu_cpg.png}
    \caption{The central pattern generator proposed by \citet{xu2018}.  Note the structure of the neurons, especially the head oscillator driving the descending pathway.}
    \label{fig: xu_cpg}
\end{figure}

@@

\begin{equation}
    \label{eq: fhn}
    \begin{aligned}
        \frac{dv}{dt} &= f(v) + w + I_\mathrm{ext}\\
        \frac{dw}{dt} &= \varepsilon(v - \gamma w + \alpha)\\
        f(v) &= \frac{v^3}{3} - v
    \end{aligned}
\end{equation}

@@

% remark that if the w and v nullclines come near each other, you see a critical slowdown.
% In Hodgkin-Huxley notation, $n$ is the potassium channel.  In the FHN formulation, this corresponds to the $w$ gate variable.
\begin{figure}[h!]
    \label{fig: nm nullclines}
    \centering
    \includegraphics{figures/neuron_dynamics/neuron_dynamics.pdf}
    % note that in the vicinity of the v-nullcline, the w-nullcline is essentially straight.
    \caption{Dynamics of \textit{C. elegans} neurons following Xu \textit{et al.} \cite{xu2018} (left) and Fitzhugh-Nagumo neurons \cite{fitzhugh1955, nagumo1962} (right).}
\end{figure}

@@

\begin{equation}
    \label{eq: fhnd}
    \begin{aligned}
        \frac{dv}{dt}   &= f(v) + w  + \color{black}{D(Δv)}\\
        \frac{dw}{dt}   &= \varepsilon (v - \gamma w + \alpha)\\
        f(v) &= \frac{v^3}{3} - v
    \end{aligned}
\end{equation}

@@

% Structure:
% - Describe the model, how the neurons are interconnected,
%   as well as how the model in general functions (as 
%   a function which solves a system of ODEs)
% - Describe how the code works, in stages:
%    - Initialize parameters
%    - Integrate the solution to the system of DEs
%    - Simulate the muscle response using smoothing
%    - Construct the worm's shape using splines with the angle approach
%
% Basically, say enough that someone could replicate the idea without seeing the code at all.

We performed the simulations in Python, using the standard SciPy ODE solvers, which wrap LSODA, the Livermore Solver for Ordinary Differential Equations. % TODO list solver here

The main function of the code solves the FitzHugh-Nagumo equations of \Fref{eq: fhnd}, having one set of such equations per neuron, which is materialized from our CPG connectome, when given the state and a set of unchanging parameters.  This procedure is followed by the ODE solver which solves the trajectory of the system given some initial conditions; for example, if the neuron does not initially fire, the system may remain in a steady state and not oscillate.

\subsection{Overview of the methods}

We are outlining the code here, but the full code can be found at *** add clean github link here ***.

@@

\subsection{Initializing the model's parameters}

% Mention that the FHN eqns are nondimensional and so all quantities have dimension but not units, i.e., arbitrary scaling.

\subsection{Integrating the solution to the system of differential equations}

\subsection{Simulating the worm's muscular response}

@@

\subsection{Constructing the worm's movement}

@@

\begin{figure}[h!]
    \centering
    \pgfplotsset{}
    % \inputtikz{figures/anal_sim_comp/anal_sim_comp}
    \includegraphics[height=9cm]{figures/anal_sim_comp/anal_sim_comp.pdf}
    \caption{Comparison between computationally simulated dynamics (top row) and observed dynamics in the analog circuit (bottom row).  (a) is the nullcline for the FitzHugh-Nagumo (simulated) equations, with the straight line being the $w$-nullcline and the cubic-like line being the $v$-nullcline; (b) is the "head oscillator" simulated (two neurons which mutually inhibit each other); (c) is the "descending pathway" simulated (one neuron driving another).  (d) is the nullcline for the analog circuit - a piecewise linear approximation to the cubic, (e) is the analog version of the head oscillator, and (f) is the analog version of the descending pathway.}
    \label{fig: anal_sim_comp}
\end{figure}
